A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2024-11-11 latex-results/01DataDescription.tex
typos fixed
config updated

The original studies included 207 participants (\(N_{S1}=\) 23,
\(N_{S2}=\) 113, \(N_{S3}=\) 71) with a total of 10,297 ESM
measurements. Each of the studies focused on recently arrived
first-generation migrants and each study included a number of
idiosyncratic variables relevant for the broader research collective.
For our empirical example, we focus on the variables that were collected
during the ESM surveys and were available in all three studies. Variable
selection and preparation are described as part of the illustration
below but for additional methodological details about the study setup
see \citet[][]{Kreienkamp2022b}.

>>>>

The original studies included 207 participants (\(N_{S1}=\) 23,
\(N_{S2}=\) 113, \(N_{S3}=\) 71) with a total of 10,297 ESM
measurements. Each of the studies focused on recently arrived
first-generation migrants and each study included a number of
idiosyncratic variables relevant for the broader research collective.
For our empirical example, we focus on the variables that were collected
during ESM surveys and were available in all three studies. Variable
selection and preparation are described as part of the illustration
below but for additional methodological details about the study setup
see \citet[][]{Kreienkamp2022b}. Each study was approved by the ethics
board of the university of origin and all participants gave informed
consent.

### 2024-11-11 latex-results/02InputVariables.tex
feedback lfb
internal feedback lfb
config updated

For our illustration, we include 12 variables that were available across
all three studies and captured information about the participant's
interactions, as well as cognitive-, emotional-, and motivational self
in relationship with the majority group. We chose these aspects in
particular because (1) the interaction-specific information exemplified
the structural missingness issue of modern ESM data and (2) the
motivational, emotional, and cognitive experience offered a diverse
conceptualization of migration experience (beyond behavioral
measurements) that is becoming more common in the literature
\citep[][]{Kreienkamp2022d}. The breadth of the included variables also
showcases the utility of the method for a growing body of literature
that considers heterogeneous and complex concepts. As a result, the
number of included variables is also on the higher end for psychological
concepts and additionally allows us to showcase the efficiency benefits
of the method and offers a reasonable use case for the optional feature
reduction step.

Once the important variables have been selected, the data needs to be
prepared for the analysis steps. Importantly, this not only means
validating and cleaning the data (e.g., re-coding, removing duplicate or
unwanted measurements) but also making the time series comparable. Two
important steps are making the time-frames and response scales
comparable across participants --- for example, by choosing a time frame
that is common to most participants and standardizing the participants'
responses \citep{liao2005}.

In our illustration data set, the studies differed substantially in the
maximum length of participation (\(\text{max}(t_{S1})=\) 63,
\(\text{max}(t_{S2})=\) 69, \(\text{max}(t_{S3})=\) 155). This was
likely due to the option to continue participation without compensation
in the later two studies. To make the three studies comparable in
participation and time frames, we iteratively removed all measurement
occasions and participants that had more than 45\% missingness
\citep[which was in line with the general rcecommendation for data that might still need to rely on imputations for later model testing][]{Madley-Dowd2019}.
This procedure let to a final sample of 157 participants, who jointly
produced 8,132 measurements. Importantly, both the participant
repsonse-patterns and the time frame were now substantially more
comparable (\(\text{max}(t_{S1})=\) 61, \(\text{max}(t_{S2})=\) 60,
\(\text{max}(t_{S3})=\) 67). Full methodological details are available
in Online Supplemental Material A, but basic item information,
descriptives, and correlations are also available in
\tblref{tab:descrLong}.

>>>>

For our illustration, we include 12 variables that were measured as part
of the ESM surveys in all three studies and captured information about
the participant's interactions, as well as the cognitive-, emotional-,
and motivational self in relationship with the majority group (see
\tblref{tab:var_selection} for an overview). We chose these aspects in
particular because (1) the interaction-specific information exemplified
the structural missingness issue of modern ESM data and (2) the
motivational, emotional, and cognitive experience offered a diverse
conceptualization of migration experience (beyond behavioral
measurements) that is becoming more common in the literature
\citep[][]{Kreienkamp2022d}. The breadth of the included variables also
showcases the utility of the method for a growing body of literature
that considers heterogeneous and complex concepts. As a result, the
number of included variables is also on the higher end for psychological
concepts and additionally allows us to showcase the efficiency benefits
of the method and offers a reasonable use case for the feature reduction
step.

\input{tables/var_selection}

Once the important variables have been selected, the data needs to be
prepared for the analysis steps. Importantly, this not only means
validating and cleaning the data (e.g., re-coding, combining scale
items) but also making the time-series comparable. Making the
time-frames and response scales comparable across participants, for
example, includes choosing a time frame that is common to most
participants
\citep['data preparation' and 'data cleaning' in \fgrref{fig:TSCFlow}; also see][]{liao2005}.

In our illustration data set, the studies differed substantially in the
maximum length of participation (\(\text{max}(t_{S1})=\) 63,
\(\text{max}(t_{S2})=\) 69, \(\text{max}(t_{S3})=\) 155). This was
likely due to the option to continue participation without compensation
in the latter study. To make the three studies comparable in
participation and time frames, we iteratively removed all measurement
occasions and participants that had more than 45\% missingness
\citep[which was in line with the general recommendation for data that might still need to rely on imputations for later model testing; see][]{Madley-Dowd2019}\footnote{Please note that for cases where the clustering is the main analysis, this high missingness threshold may be too conservative. As part of our validation analyses in \appref[]{app:ValidationAnalyses} we compare the model presented here with varying levels of missing data allowed.}.
This procedure led to a final sample of 157 participants, who jointly
produced 8,132 survey responses. Importantly, both the participant
response patterns and the time frame were now substantially more
comparable (number of measurement occasions per person: \(t_{S1}\): min
= 40, max = 61, mean = 57.33, sd = 4.69; \(t_{S2}\): min = 33, max = 60,
mean = 49.05, sd = 6.73; \(t_{S3}\): min = 36, max = 65, mean = 54.20,
sd = 7.04). It is important to consider that some time series features
may be less reliable when the number of measurement occasions per person
is low (e.g., below 30 measurements per person), and this should be
taken into account when conducting similar analyses. Full methodological
details are available in Online Supplemental Material A, but basic item
information, descriptives, and correlations are also available in
\tblref{tab:descrLong}.

### 2024-11-11 latex-results/03FeatureExtraction.tex
typo fixed
config updated

For the central tendency feature of our illustration we chose the robust
\textit{median}, which can avoid potential issues with non-normally
distributed time series responses or outliers \citep{weisberg1992}. To
calculate the \textit{median} (\(M\)), we let \(X_{ij}\) be the ordered
list of values from the time series of variable \(j\) for participant
\(i\). The calculation depends on whether the number of measurements in
a time series \(n\) is odd or even.

\begin{equation} \label{eq:median}
  M(X_{ij}) = 
    \begin{cases}
      X \left[ \frac{n+1}{2} \right] & \text{if $n$ is odd} \\
      \frac{X \left[ \frac{n}{2} \right] + X \left[ \frac{n}{2} +1 \right]}{2} & \text{if $n$ is even}
    \end{cases}
\end{equation}

@@

Variability captures the degree to which a set of data differs from the
central tendency and is sometimes also referred to as the dispersion or
spread of the data \citep{weisberg1992}. In time series analyses,
variability is conceptually important because information about the
distribution and diversity of the data has been found to be indicative
of worse psychological states \citep{myin-germeys2018, helmich2021}.
Person-level differences of ESM measurements have, for example, been
associated with higher levels of psycho-pathological recurrences among
depression patients \citep{timm2017}. As such, psychological researchers
and practitioners are often empirically interested in between-person
differences in variability. Researchers on polarization and
radicalization might for example ask: ``Are people settled in their
attitudes towards migrants or do they vary across the measurement
period?''

For our illustration data, we chose to capture the time series
variability with the \textit{Median Absolute Deviation} (\(MAD\)), where
we calculate the \textit{median} (\(M\); calculated as in
\equatref{eq:median}) for the absolute deviations of measurement \(x\)
at time point \(t\) for participant \(i\) and variable \(j\) from the
median of that time series \(X\). We again chose the robust statistic
because the Median-based measure is less affected by non-normal
distributions and extreme values or outliers compared to other measures
of variability like the standard deviation \citep{weisberg1992}

\begin{equation} \label{eq:mad}
  MAD(X_{ij}) = M(\left| x_{ijt} - M(X_{ij}) \right|)
\end{equation}

@@

Instability captures the average change between two consecutive
measurements \citep{ebner-priemer2009}. While instability is
conceptually related to the variability feature, variability does not
take into account temporal dependency, whereas instability looks at the
`jumpy-ness' of the data over time. In other words, variability reflects
the range or diversity of values in a time series data, while
instability reflects the fluctuation or inconsistency in a time series
data over time \citep{trull2008}. For example, if a person has rapid and
extreme changes in mood their mood is highly unstable, while if a
person's mood responses span a wide range over the entire study period,
their mood is highly variable \citep{jahng2008}. Within psychological
time series, instability measurements have especially been important in
the research of borderline personality disorder \citep{trull2008} and
suicidality \citep{kivela2022}, but also in understanding early warning
signals more generally \citep{wichers2019}. Conceptually, the
instability feature, thus, relates to a broad range of research
questions, including: ``What is the nature of the identification changes
in those who start working in a new country?'' or ``Do strong daily
fluctuations in self-esteem reflect the process of identity formation in
adolescents?''

@@

Another common measurement of instability is the
\textit{Mean of the Squared Successive Differences} (\(MSSD\)), which is
often preferred where differences in magnitude are more important than
the frequency of those changes, for example, when big shifts in time
series are considered more impactful or when outliers are meaningful and
need to be taken into account \citep{chatfield2003}.

\paragraph{Self-similarity.}

Self-similarity in time series data refers to the property of a time
series to exhibit similar patterns of behavior over different time
scales \citep{dmello2021}. That is, self-similarity describes how much a
measurement carries over to future measurements. One important
self-similarity in psychological time series is \texit{inertia} --- how
much a measurement carries over to its next measurement
\citep{kuppens2010, suls1998}. If inertia is high a development tends to
stay in a certain state. Because high inertia is resistant to change, in
emotion dynamics high inertia of negative affect has been found to be
indicative of under-reactive systems and to be characteristic of
psychological maladjustment \citep{kuppens2010}. In a similar vein, high
inertia in negative affect at baseline was even predictive of the
initial onset of depression \citep{kuppens2012}. Conceptually, inertia
is more broadly connected to research questions such as:
\texttt{Do\ patients\ stay\ in\ a\ negative\ mood\ for\ several\ measurements?\textquotesingle{}\textquotesingle{}\ or}Do
migrants stay with their language practice for several days at a
time?'\,'

@@

@@

Changes in psychology are not always linear, instead, nonlinearity is a
common feature of psychological time series \citep{hayes2007}. As an
example, episodic disorders, such as depression, are most likely best
described as non-linear systems \citep{hosenfeld2015}. Similarly,
patients in recovery from depression showed sudden changes in the
improvement of depression \citep{helmich2020a}. But also substance abuse
\citep{boker1998} or attitude changes rarely develop linearly
\citep{vandermaas2003}. Conceptually, researchers might have research
questions about the type of the development: ``Is the development of
well-being a nonlinear process?'' as well as the shape and structure of
the development: ``How many spikes in well-being did a migrant
experience?''

We summarized the nonlinear trend with the
\textit{estimated degrees of freedom} of an empty GAM spline model. The
\(edf\) summarizes the \textit{wiggliness} of a spline trend line
\citep{wood2017}. The degrees of freedom of a spline model are primarily
determined by the number of knots and the order of the spline. For
instance, a cubic spline with \(k\) knots has \(k\)+3 degrees of freedom
\citep{faraway2016}. However, in a penalized spline framework, which is
commonly used for GAMs, the effective degrees of freedom can be less
than \(k\)+3. This is because the model employs a smoothing parameter to
control the trade-off between the complexity (flexibility) of the model
and its fit to the data, thereby penalizing overly complex models and
potentially reducing the effective degrees of freedom \citep{marx1998}.
Intuitively then an edf of 1 would be equivalent to a linear
relationship (i.e., one linear slope parameter), whereas a higher edf
(particularly an edf \textgreater{} 2) is indicative of a non-linear
trend. The estimated degrees of freedom are commonly based on a concept
called `effective degrees of freedom' and can be represented as the
trace \(tr\) (i.e., the sum of the diagonal elements) of the smoother
matrix \(S\), a symmetric matrix that maps from the raw data to the
smooth estimates \citep{wood2017}.

@@

Beyond our main features of interest, we also extracted the
participant's number of completed ESM measurements to ensure that the
clusters are comparable in that regard (i.e., to exclude spurious
explanations for the cluster assignments). After the feature extraction,
we found that about 1.40\% of the extracted features are missing across
the 72 features per participant. This might, for example, happen if
participants do not have two subsequent measurements with outgroup
interactions, so that an autocorrelation with lag-1 cannot be calculated
for the contact-specific variables. The small number of missing values
indicates that the feature-based approach indeed largely avoids the
structural missingness issue. However, even the few missing values can
be an issue for some feature reduction or feature clustering algorithms.
We, thus, impute the missing feature values with a single predictive
mean matching imputation using the MICE library \citep[][]{buuren2011}.
Note again that with this procedure we only need to impute an extremely
small number of missing values as most feature calculations can use the
available data instead.

>>>>

For the central tendency feature of our illustration, we chose the
median (\(M\)), which effectively addresses potential complications
arising from non-normally distributed responses or outliers within time
series datasets \citep{weisberg1992}. To compute the median, it is
imperative to differentiate between two types of time series
representations for a given variable \(j\) related to participant \(i\):
the chronological series and the ordered series. The chronological time
series, denoted by \(X_{ij}\), encapsulates the sequence of observations
\(\{x_{ij1}, x_{ij2}, ..., x_{ijn}\}\) for variable \(j\) concerning
participant \(i\), organized by their temporal occurrence. Here,
\(x_{ijt}\) signifies a specific observation at time \(t\) within this
sequence. In contrast, the ordered time series, represented as
\(\mathbf{X}_{ij}\), is derived from \(X_{ij}\) by sorting the
observations in ascending order of magnitude. This ordered set is
expressed as
\(\{\mathbf{x}_{ij1}, \mathbf{x}_{ij2}, ..., \mathbf{x}_{ijn}\}\), with
each \(\mathbf{x}_{ijk}\) corresponding to the \(k\)-th element in the
reordered series \(\mathbf{X}_{ij}\).

The median \(M(\mathbf{X}_{ij})\) is then the value located precisely at
the center of the ordered time series \(\mathbf{X}_{ij}\). Depending on
whether the total number of observations (\(n\)) is odd or even, the
median is either the middel \(k\)-th element if \(n\) is odd, or the
average of the two middle values if \(n\) is even:

\begin{equation} \label{eq:median}
  M(\mathbf{X}_{ij}) = 
  \begin{cases}
    \mathbf{x}_{ij\left(\frac{n+1}{2}\right)} & \text{if $n$ is odd} \\
    \frac{\mathbf{x}_{ij\left(\frac{n}{2}\right)} + \mathbf{x}_{ij\left(\frac{n}{2} + 1\right)}}{2} & \text{if $n$ is even}
  \end{cases}
\end{equation}

This approach ensures that the median is a reliable indicator of central
tendency in time series analysis, unaffected by data distribution
asymmetries or the presence of outliers.

@@

Variability captures the degree to which a set of data differs from the
central tendency and is sometimes also referred to as the dispersion or
spread of the data \citep{weisberg1992}. Common measurements of the
variability are the variance or standard deviation as well as their
robust counter parts. In time series analyses, variability is
conceptually important because information about the distribution and
diversity of the data has been found to be indicative of worse
psychological states \citep{myin-germeys2018, helmich2021}. Person-level
differences of ESM measurements have, for example, been associated with
higher levels of psycho-pathological recurrences among depression
patients \citep{timm2017}. As such, psychological researchers and
practitioners are often empirically interested in between-person
differences in variability. Researchers on polarization and
radicalization might for example ask: ``Are people settled in their
attitudes towards migrants or do they vary across the measurement
period?''

For our illustration data, we chose the
\textit{Median Absolute Deviation} (\(MAD\)) to gauge the variability
within our time series data. This choice is motivated by the robustness
of MAD, particularly its resilience to the effects of non-normal
distributions and outliers, which can significantly skew traditional
variability measures like the standard deviation \citep{weisberg1992}.
For a given variable \(j\) and participant \(i\), the MAD is calculated
by first determining the median (\(M\)) of the ordered time series
\(\mathbf{X}_{ij}\) as outlined in \equatref{eq:median}. We then compute
the absolute deviations of each observation in the time series
\(X_{ij}\) from this median value. Specifically, for each time point
\(t\), we calculate the absolute difference between \(x_{ijt}\) and the
series median \(M(\mathbf{X}_{ij})\). The MAD is then the median of
these absolute deviations:

\begin{subequations}\label{eq:mad}
    \begin{align} 
      MAD(X_{ij}) &= M(\left| x_{ijt} - M(\mathbf{X}_{ij}) \right|) \label{eq:mad_general} \\
                  &= M(\{ |x_{ij1} - M(\mathbf{X}_{ij})|, |x_{ij2} - M(\mathbf{X}_{ij})|, \ldots, |x_{ijn} - M(\mathbf{X}_{ij})| \}) \label{eq:mad_detailed}
    \end{align}
\end{subequations}

The calculation of MAD focuses on the magnitudes of deviations, ensuring
it provides a robust measure of dispersion that reflects the inherent
variability in the time series data.

@@

Instability captures the average change between two consecutive
measurements \citep{ebner-priemer2009, jahng2008}. While instability is
conceptually related to the variability feature, variability does not
take into account temporal dependency, whereas instability looks at the
`jumpy-ness' of the data over time. In other words, variability reflects
the range or diversity of values in the un-ordered time series data,
while instability reflects the fluctuation or inconsistency in a time
series data over time \citep{trull2008, houben2015, koval2013}. For
example, if a person has rapid and extreme changes in mood their mood is
highly unstable, while if a person's mood responses span a wide range
over the entire study period, their mood is highly variable
\citep[note that this does not need to be rapidly changing or instable, e.g., when there is linear increase over time; also see][]{jahng2008}.
Within psychological time series, instability measurements have
especially been important in the research of borderline personality
disorder \citep{trull2008} and suicidality \citep{kivela2022}, but also
in understanding early warning signals more generally
\citep{wichers2019}. Conceptually, the instability feature, thus,
relates to a broad range of research questions, including: ``What is the
nature of the identification changes in those who start working in a new
country?'' or ``Do strong daily fluctuations in self-esteem reflect the
process of identity formation in adolescents?''

@@

Another common measurement of instability is the
\textit{Mean of the Squared Successive Differences} (\(MSSD\)), which is
often preferred where differences in magnitude are more important than
the frequency of those changes, for example, when big shifts in time
series are considered more impactful or when outliers are meaningful and
need to be taken into account \citep{chatfield2003, bos2019}. For
psychological ESM data, some research suggests that amplitude and
frequency might predict different health outcomes and can be
investigated jointly \citep{wang2012a, jahng2008}.

\paragraph{Temporal dependence.}

Temporal dependence in time series data refers to the degree to which a
time series is influenced by its past values, exhibiting patterns of
behavior that may be regular over different time scales
\citep{dmello2021}. In the context of psychological time series, an
important aspect of temporal dependence is \textit{inertia} --- how much
a measurement carries over to its next measurement
\citep{kuppens2010, suls1998}. If inertia is high a development tends to
stay in a certain state. Because high inertia is resistant to change, in
emotion dynamics high inertia of negative affect has been found to be
indicative of under-reactive systems and to be characteristic of
psychological maladjustment \citep{kuppens2010}. In a similar vein, high
inertia in negative affect at baseline was predictive of the initial
onset of depression \citep{kuppens2012}. Conceptually, inertia is more
broadly connected to research questions such as: ``Do patients stay in a
negative mood for several measurements?'' or ``Do migrants stay with
their language practice for several days at a time?''

@@

While inertia captures the simplest case of temporal dynamics, lag-1, we
acknowledge that temporal dependence in psychological time series may
also exhibit more complex relationships, including higher lagged auto
correlations or cyclical relationships (fourier coefficients, or
continuous wavelet transforms are often used to capture such
relationships).

@@

Changes in psychology are not always linear, instead, nonlinearity is a
common feature of psychological time series \citep{hayes2007}. As an
example, episodic disorders, such as depression, are often best
described as non-linear systems \citep{hosenfeld2015}. Similarly,
patients in recovery from depression showed sudden changes in the
improvement of depression \citep{helmich2020a}. But also substance abuse
\citep{boker1998} or attitude changes rarely develop linearly
\citep{vandermaas2003}. Conceptually, researchers might have research
questions about the type of the development: ``Is the development of
well-being a nonlinear process?'' as well as the shape and structure of
the development: ``How many spikes in well-being did a migrant
experience?''

We summarized the nonlinear trend with the
\textit{estimated degrees of freedom} of an empty GAM spline model. The
\(edf\) summarizes the \textit{wiggliness} of a spline trend line
\citep{wood2017, bringmann2017}. The degrees of freedom of a spline
model are primarily determined by the number of knots and the order of
the spline. For instance, a cubic spline with \(k\) knots has \(k\)+3
degrees of freedom
\citep{faraway2016, haslbeck2021a, castro-alvarez2024}. However, in a
penalized spline framework, which is commonly used for GAMs, the
effective degrees of freedom can be less than \(k\)+3. This is because
the model employs a smoothing parameter to control the trade-off between
the complexity (flexibility) of the model and its fit to the data,
thereby penalizing overly complex models and potentially reducing the
effective degrees of freedom \citep{marx1998}. Intuitively then an edf
of 1 would be equivalent to a linear relationship (i.e., one linear
slope parameter), whereas a higher edf (particularly an edf
\textgreater{} 2) is indicative of a non-linear trend. The estimated
degrees of freedom are commonly based on a concept called `effective
degrees of freedom' and can be represented as the trace \(tr\) (i.e.,
the sum of the diagonal elements) of the smoother matrix \(S\), a
symmetric matrix that maps from the raw data to the smooth estimates
\citep{wood2017}.

@@

\paragraph{Additional considerations.}

Beyond our main features of interest, we also extracted the
participant's number of completed ESM measurements to ensure that the
clusters are comparable in that regard (i.e., to exclude spurious
explanations for the cluster assignments). After the feature extraction,
we found that about 1.40\% of the extracted features are missing across
the 72 features per participant. This might, for example, happen if
participants do not have two subsequent measurements with outgroup
interactions, so that an autocorrelation with lag-1 cannot be calculated
for the contact-specific variables. The small number of missing values
indicates that the feature-based approach indeed largely avoids the
structural missingness issue. However, even the few missing values can
be an issue for some feature reduction or feature clustering algorithms.
We, thus, impute missing feature values via predictive mean matching
(PMM) with the MICE package in R, employing a single imputation and
specifying a maximum of 50 iterations and a fixed seed for convergence
and reproducibility \citep[][]{buuren2011}. We chose PMM for its ability
to preserve the original data distribution without assuming normality
and robustly handling multiple data types \citep{vanbuuren2006}. Note
again that with this procedure we only need to impute an extremely small
number of missing values as most feature calculations can use the
available data instead.

### 2024-11-11 latex-results/04FeatureReduction.tex
config updated

For our own illustration data, we chose a feature projection method to
reduce the dimensionality of our extracted features. We particularly
chose the feature projection method for its broad applicability. We,
specifically, selected the commonly used
\textit{principal component analysis} (PCA). Some of the more
tailor-made feature selection algorithms can be more accurate in
reducing the feature dimensionality and might retain feature importance
information more directly, depending on the specific data structure.
However, PCAs have the distinct benefit that they are well-established
within the psychometric literature \citep{jolliffe2011} and can broadly
be applied to a wide variety of studies in an automatized manner
\citep{abdi2010}. As our aim is to present a general illustration that
can also be adopted across use cases, we present the workflow using a
PCA here but we encourage users to consider more specialized methods as
well.

To use the PCA with our extracted time series features, we first
standardize all features across participants to ensure that all features
are weighted equally. We then enter all 72 features into the analysis.
The PCA uses linear transformations in such a way that the first
component captures the most possible variance of the original data
\citep[e.g., by finding a vector that maximizes the sum of squared distances][]{jolliffe2002, abdi2010}.
The following components will then use the same method to iteratively
explain the most of the remaining variance while also ensuring that the
components are linearly uncorrelated \citep{shlens2014}. In practice,
this meant that the PCA decomposed the 72 features into 72 principal
components but now (because of the uncorrelated linear transformations)
the first few principal components will capture a majority of the
variance. We can then decide how much information (i.e., variance) we
are willing to sacrifice for a reduced dimensionality. A common rule of
thumb is to use the principal components that jointly explain 70--90\%
of the original variance
\citep[i.e., cumulative percentage explained variance; e.g.,][]{jackson2003}.
For our illustration, we select the first 27 principal components that
explain 80\% of the variance in the original 72 features (reducing the
dimensionality by 62.50\%). For the extracted principal components we
save the 27 principal component scores for each participant (i.e., the
participants' coordinates in the reduced dimensional space; PC-scores).

We would like to comment on two practical matters when using principal
components --- the amount of dimensionality reduction and the
interpretation of the principal components. As for the expected
dimensionality reduction, given its methodology, PCAs tend to `work
better' at reducing dimensions with (highly) correlated variables
\citep[e.g.,][]{jolliffe2002}. Thus, with a set of very homogeneous
variables and features users will need fewer principal components to
explain a large amount of variance, while a more diverse set of
variables and features will tend to require more principal components to
capture the same amount of variance \citep[e.g.,][]{abdi2010}. Our 27
principal components are still a relatively high number of variables but
this is not surprising as we chose a diverse conceptualization and a
diverse set of time series features. As for interpretability, PCA allows
users to extract information on the meaning of the principal components.
In particular, because the principal components are linear combinations
of the original features, users can extract the relative importance of
each feature for the extracted principal components (i.e., the
eigenvectors). While this can be useful in understanding the variance in
the original data or help with manual feature selection, we use the PCA
purely to reduce the dimensionality for the clustering step. Instead of
relying on the principal components, we use the original features of
interest to interpret the later extracted clusters. We particularly
advocate for such an approach if all original features are considered
meaningful in understanding the time series and users would like to
retain the features for interpretation (irrespective of the features'
importance).

>>>>

For our own illustration data, we chose a feature projection method to
reduce the dimensionality of our extracted features. We particularly
chose the feature projection method for its broad applicability. We,
specifically, selected the commonly used
\textit{principal component analysis} (PCA). Some of the more
tailor-made feature selection algorithms can be more accurate in
reducing the feature dimensionality and might retain feature importance
information more directly, depending on the specific data structure.
However, PCAs have the distinct benefit that they are well-established
within the psychometric literature \citep{jolliffe2011} and can broadly
be applied to a wide variety of studies in an automatized manner
\citep{abdi2010}. As our aim is to present a general illustration that
can also be adopted across use cases, we present the workflow using a
PCA here but we encourage users to consider more specialized methods as
well (we provide an example decision guide in \sicontext).

To use the PCA with our extracted time series features, we first
standardize all features across participants to ensure that all features
are weighted equally \citep[][]{horne2020}. We then enter all 72
features into the analysis. The PCA uses linear transformations in such
a way that the first component captures the most possible variance of
the original data
\citep[e.g., by finding a vector that maximizes the sum of squared distances][]{jolliffe2002, abdi2010}.
The following components will then use the same method to iteratively
explain the most of the remaining variance while also ensuring that the
components are linearly uncorrelated \citep{shlens2014}. In practice,
this meant that the PCA decomposed the 72 features into 72 principal
components but now (because of the uncorrelated linear transformations)
the first few principal components will capture a majority of the
variance. We can then decide how much information (i.e., variance) we
are willing to sacrifice for a reduced dimensionality. A common rule of
thumb is to use the principal components that jointly explain 70--90\%
of the original variance
\citep[i.e., cumulative percentage explained variance; e.g.,][]{jackson2003}.
For our illustration, we select the first 27 principal components that
explain 80\% of the variance in the original 72 features (reducing the
dimensionality by 62.50\%). For the extracted principal components we
save the 27 principal component scores for each participant (i.e., the
participants' coordinates in the reduced dimensional space; PC-scores).

We would like to comment on two practical matters when using principal
components --- the amount of dimensionality reduction and the
interpretation of the principal components. As for the expected
dimensionality reduction, given its methodology, PCAs tend to `work
better' at reducing dimensions with (highly) correlated variables
\citep[e.g.,][]{jolliffe2002}. Thus, with a set of very homogeneous
variables and features users will need fewer principal components to
explain a large amount of variance, while a more diverse set of
variables and features will tend to require more principal components to
capture the same amount of variance \citep[e.g.,][]{abdi2010}. Our 27
principal components are still a relatively high number of variables but
this is not surprising as we chose a diverse conceptualization and a
diverse set of time series features. As for interpretability, PCA allows
users to extract information on the meaning of the principal components.
In particular, because the principal components are linear combinations
of the original features, users can extract the relative importance of
each feature for the extracted principal components (i.e., the
eigenvectors). While this can be useful in understanding the variance in
the original data or help with manual feature selection, we use the PCA
here purely to reduce the dimensionality for the clustering step.
Instead of relying on the principal components, we use the original
features of interest to interpret the later extracted clusters. We
particularly advocate for such an approach if all original features are
considered meaningful in understanding the time series and users would
like to retain the features for interpretation (irrespective of the
features' importance).

### 2024-11-11 latex-results/06ClusterEvaluationPerformance.tex
config updated

In our own illustration example, we used the \texttt{cluster.stats()}
function from the \texttt{fpc} \textsf{R} package, which calculates a
wide variety of internal cluster validity statistics for each of the
extracted clustering solutions. With real-world data no single
evaluation measure is likely perfect, and different measures may produce
different results depending on the characteristics of the data and the
research question being addressed \citep{kittler1998}. It is therefore
important to consider a variety of evaluation measures and to carefully
interpret the results in the context of the specific analysis
\citep{vinh2009}. We found that across most indices, the analysis with
\(k=2\) clusters performed the best. Three commonly reported indices we
would like to highlight are the comparison of within clusters sum of
squares, the average silhouette score, and the Calinski-Harabasz index.
The first statistic we looked at was the total within-cluster sum of
square \(WCSS\) (see also \equatref{eq:kWCSS}). While the within-cluster
variation will naturally decrease with (more) smaller clusters, we
observed that the decrease in \(WCSS\) was largest until \(k=2\) after
which the decrease was much smaller. This method is also sometimes
referred to as the `elbow method' \citep{syakur2018}. We then looked at
a second, commonly used measure, the average silhouette score. This
statistic measures the degree to which each time feature data point is
similar to other points within the same cluster, compared to points in
other clusters \citep{rousseeuw1987}. In our case, the \(k=2\) solution
maximized the silhouette coefficient (\(s_2=\) 0.09). Finally, the
Calinski-Harabasz index assesses the compactness and separation of the
clusters by assessing the ratio of the sum of between-clusters
dispersion and of inter-cluster dispersion for all clusters --- thus,
the higher the score the better the performances \citep{calinski1974}.
In our case, the \(k=2\) solution also showed the highest
Calinski-Harabasz index (\(CH_2=\) 16.38; a full table of all extracted
validity statistics is available in Supplemental Material
A)\footnote{It is important to note that another commonly assessed aspect of the evaluation is determining the stability and robustness of the clusters \citep{berkhin2006}. This can be assessed by evaluating the sensitivity of the clusters to different feature sets or clustering algorithms, or by using techniques such as bootstrapping to assess the uncertainty of the clusters \citep{vinh2009}. Especially when comparing different clustering algorithms one common index is the Bayesian information criterion (BIC), where a lower BIC indicates that a model is more representative of the data \citep{vandeschoot2017}.}.
In the final \(k=2\) solution the k-means analysis also assigned a
relatively even number of participants to cluster 1 (\(n_{C_1}=\) 76)
and cluster 2 (\(n_{C_1}=\) 80).

>>>>

In our own illustration example, we used the \texttt{cluster.stats()}
function from the \texttt{fpc} \textsf{R} package, which calculates a
wide variety of internal cluster validity statistics for each of the
extracted clustering solutions. With real-world data, no single
evaluation measure is likely to be perfect. Different measures may yield
varying results based on the data characteristics and the research
question at hand \citep{kittler1998}. It is therefore important to
consider a variety of evaluation measures and to carefully interpret the
results in the context of the specific analysis \citep{vinh2009}. We
found that across most indices, the analysis with \(k=2\) clusters
performed the best. Three commonly reported indices we would like to
highlight are the comparison of within clusters sum of squares, the
average silhouette score, and the Calinski-Harabasz index. The first
statistic we looked at was the total within-cluster sum of square
\(WCSS\) (see also \equatref{eq:kWCSS}). While the within-cluster
variation will naturally decrease with (more) smaller clusters, we
observed that the decrease in \(WCSS\) was largest until \(k=2\), after
which the decrease was much smaller. This method is also sometimes
referred to as the `elbow method' \citep{syakur2018}. We then looked at
a second, commonly used measure, the average silhouette score. This
statistic measures the degree to which each time feature data point is
similar to other points within the same cluster, compared to points in
other clusters \citep{rousseeuw1987}. In our case, the \(k=2\) solution
maximized the silhouette coefficient (\(s_2=\) 0.09). Finally, the
Calinski-Harabasz index assesses the compactness and separation of the
clusters by assessing the ratio of the sum of between-clusters
dispersion and of inter-cluster dispersion for all clusters --- thus,
the higher the score the better the performances \citep{calinski1974}.
In our case, the \(k=2\) solution also showed the highest
Calinski-Harabasz index (\(CH_2=\) 16.38; a full table of all extracted
validity statistics is available in Supplemental Material
A)\footnote{It is important to note that another commonly assessed aspect of the evaluation is determining the stability and robustness of the clusters \citep{berkhin2006}. This can be assessed by evaluating the sensitivity of the clusters to different feature sets or clustering algorithms, or by using techniques such as bootstrapping to assess the uncertainty of the clusters \citep{vinh2009}. Especially when comparing different clustering algorithms one common index is the Bayesian information criterion (BIC), where a lower BIC indicates that a model is more representative of the data \citep{vandeschoot2017}.}.
In the final \(k=2\) solution the k-means analysis also assigned a
relatively even number of participants to cluster 1 (\(n_{C_1}=\) 76)
and cluster 2 (\(n_{C_1}=\) 80).

To ensure that the clustering is necessary in the first place, we also
compare the performance to a single cluster solution (i.e., a single
centroid). The comparison to this \(k=1\) solution is slightly different
because metrics like the between-cluster separation are not available.
Nonetheless, comparing the within-cluster sums of squares (SS) and the
explained variance, we find that two clusters indeed outperform a single
cluster solution. Specifically, the total within-cluster SS decreased
from 8940.21 for one cluster to 8080.67 for two clusters. Additionally,
the variance explained increased from \textless.001 to 0.096 when the
cluster count was increased to two
\citep[e.g., ][; also see \situtorial\ for full results]{beijers2022}.
